msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-12 paper.tex
fixups

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
representation of the coalescent with recombination stochastic process. This
formulation was complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of the recombination. In Griffiths' ARG formulation, we
regard a realisation of the coalescent with recombination as a graph in which
nodes represent common ancestor or recombination events, and edges represent
lineages through which ancestral material flows. We may consider the ``big''
ARG, in which we wait an exponentially large time until all sampled lineages
coalescence into a single ancestor~\citep{ethier1990two}, or the ``little'' ARG
in which we wait until all marginal trees have fully coalescenced (like
Hudson's algorithm).

% Seems somewhat unfair to single out Iain and Aylwyn here, but a systematic
% review would be long and tedious, and I'm sure they won't be too upset about
% it.
However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean a representation of genetic ancestry as a graph; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material. Such
imprecision is unfortunate, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little to
clarify. As we shall see, there are situations in which the precise information
stored of the graph is important, and conflating a high-level
discussion of genetic ancestry with this mathematical object is not
helpful. One may wonder if a more generic term such as ``genetic pedigree''
would be a more useful description, in such cases where the precise form of the
encoding is not important.

@@

Although the tree sequences output by \msprime\ contain full information
about the genealogical trees, their correlation structure along the genome,
and the ancestral haplotypes on which coalescences occured, information
is lost in this mapping down from ARG space. In particular, we lose
information about ancestral haplotypes that were common ancestors but
in which no coalescences occured, and also information about the precise time
and genomic location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the compete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence.
Fig.~\ref{fig-recombination-perf}B shows that this extra ARG information
quickly comes to dominate the output as we increase the sequence length,
and the resulting tree sequence files are many times larger
than the standard minimal tree sequence representation.

>>>>

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
representation of the coalescent with recombination stochastic process. This
formulation was complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation, we
regard a realisation of the coalescent with recombination as a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages through which ancestral material flows. We may consider the ``big''
ARG, in which we wait an exponentially large time until all sampled lineages
coalescence into a single ancestor~\citep{ethier1990two}, or the ``little'' ARG
in which we wait until all marginal trees have fully coalescenced (like
Hudson's algorithm).

% Seems somewhat unfair to single out Iain and Aylwyn here, but a systematic
% review would be long and tedious, and I'm sure they won't be too upset about
% it.
However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean a representation of genetic ancestry as a graph; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material. Such
imprecision is unfortunate, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little to
clarify. As we shall see, there are situations in which the precise information
stored in the graph is important, and conflating a high-level
discussion of genetic ancestry with this mathematical object is not
helpful. One may wonder if a more generic term such as ``genetic pedigree''
would be a more useful description, in such cases where the precise form of the
encoding is not important.

@@

Although the tree sequences output by \msprime\ contain full information
about the genealogical trees, their correlation structure along the genome,
and the ancestral haplotypes on which coalescences occured, information
is lost in this mapping down from ARG space. In particular, we lose
information about ancestral haplotypes that were common ancestors but
in which no coalescences occured, and also information about the precise time
and genomic location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the complete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence.
Fig.~\ref{fig-recombination-perf}B shows that this extra ARG information
quickly comes to dominate the output as we increase the sequence length,
and the resulting tree sequence files are many times larger
than the standard minimal tree sequence representation.
